### 2017-05-30 distr-03-design.tex
Tiny textual fixes

\subsection{Pipelining of computation and loading $\pi$}

The distributed design features two instances of pipeline parallelism. First,
at the process level, sampling of the minibatch by the master is decoupled
from \textit{update\_phi} by the workers, so the master computes the minibatch
for the next iteration while the workers compute \textit{update\_phi} for the
current iteration. Second, at the threads level within the workers, loading of
$\pi$ and \textit{update\_phi} are split up into chunks, and the calculation
in \textit{update\_phi} for the current chunk is done simultaneously with
loading $\pi$ for the next chunk. The rationale for implementing pipeline
parallelism in these places of the program, is that loading $\pi$ is the
dominant contribution within the dominant stage \textit{update\_phi}, as
will be shown in Section~\ref{eval-pipeline}.  Since loading $\pi$ is the
performance bottleneck of the distributed algorithm, performing other work
in parallel shortens the critical path.

>>>>

\subsection{Pipelining of computation and loading $\pi$}
\label{sec-pipelining}

The distributed design features two instances of pipeline parallelism. First,
at the process level, sampling of the minibatch by the master is decoupled
from \textit{update\_phi} by the workers, so the master computes the minibatch
for the next iteration while the workers compute \textit{update\_phi} for the
current iteration. Second, at the threads level within the workers, loading of
$\pi$ and \textit{update\_phi} are split up into chunks, and the calculation
in \textit{update\_phi} for the current chunk is done simultaneously with
loading $\pi$ for the next chunk. The rationale for implementing pipeline
parallelism in these places of the program is that loading $\pi$ is the
dominant contribution within the dominant stage \textit{update\_phi}, as
will be shown in Section~\ref{eval-pipeline}.  Since loading $\pi$ is the
performance bottleneck of the distributed algorithm, performing other work
in parallel shortens the critical path.

### 2017-05-30 gpu-04-03-accelerator.tex
Tiny textual fixes

A key structural change applied to the algorithm is the lossless compression of
its state. This modification enabled the algorithm to process larger data sets
while maintaining all program state in memory. Moreover, as the algorithm is
data-intensive, a reduction of the state is accompanied by a decrease in its
data-intensity.
%
The data structures that occupy most memory are $\pi$ and~$\phi$, both of
which are matrices of dimension ${N}\times{K}$. However, storing both
is redundant as $\pi$ is a row-normalized copy of~$\phi$, see
Table~\ref{tbl-symbols} in Section~\ref{sec-background}. The full storage for
$\phi$ is discarded; $\phi_{ik}$ values are calculated as
$\pi_{ik}/\phi^{sum}_i$, which requires maintenance of a vector
$\phi^{sum}$ of size $N$.
Moreover, the calculation for $\phi_{ik}$ can be cached usefully.
%
The $\phi$ matrix is required in two kernels only, namely, \textit{update\_phi} and
\textit{update\_pi}. Both kernels access $\phi_i$ only for vertices~$i$ in the
mini-batch~\Minibatch, so for each iteration, the calculated values for $\phi$ are cached
in a smaller temporary matrix of size
$M\times{K}$.
%
% In this context, the initialization of the algorithm's state is performed by
% normalizing the full $\pi$ matrix in-place without the need for an additional
% matrix. Moreover, the normalization denominator of each row is cached in
% $\phi^{sum}_{i}$.
%
% Thus each read-only access to $\phi_{i,j}$ can be substituted with
% $\pi_{i,j}\phi_{i}^{sum}$. Given that the value of $\phi_{i,j}$ is consistently
% used in conjunction with $\pi_{i,j}$ and that $\phi_{i}^{sum}$ is read once and
% reused multiple times, this
This transformation trades memory storage and
bandwidth for a minimal computation overhead.
%
% On the other hand, the newly computed $\phi$ values can be stored in the
% temporary matrix and later be normalized into their corresponding rows of
% $\pi$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Single floating point vs double
Another modification to the original algorithm is migrating all of its
computation from double to single precision floating-point, analogous to the change
for the baseline implementation in Section~\ref{sec-seq-performance}.
%
This change affects the algorithm in multiple ways. For instance, on some GPUs it
reduces
the computation intensity besides the data intensity. It again halves the memory
footprint of the algorithm and frees registers.

@@

\subsubsection{Edge-centric and Vertex-centric kernels}
The work decomposition of the kernels ensures that each thread performs
independent computation in order to avoid expensive synchronization operations.
Edge-centric kernels, which operate over mini-batch edges, perform computations
over every edge in parallel.  Similarly, vertex-centric kernels that operate
over vertices in the mini-batch exploit parallelism across the selected
vertices.
% vectorization
Additionally, the kernels were vectorized to decrease instruction overhead.

\subsubsection{Memory Hierarchy Optimizations}
\label{gpu-design}
% The GPU implementation builds on top of the CPU work decomposition scheme.
% However, instead of having every thread perform independent computations, each
% block of threads shares the work associated with the single edge or vertex for
% edge-centric and vertex-centric kernels.

Since all of the kernels are data-bound, several memory organization strategies
of varying complexities were investigated to speedup the algorithm by
exploiting the GPU's memory hierarchy.
% Update_phi
As a case study, a discussion of the \textit{update\_phi} kernel is provided as it is
the predominant component of the algorithm.

@@

% Naive strategy
\paragraph*{\textbf{Naive strategy}} simply allocate temporary vectors in thread
local memory to store $Probs$ and $Grads$ which physically reside in
device memory. All memory accesses are coalesced to achieve the highest
possible bandwidth to and from device memory.

@@

% Register and shared memory using code generation and loop unrolling 
\paragraph*{\textbf{Code generation strategy}} dynamically generate the code of the kernel to
custom tailor its properties. For example, this strategy can control whether a
vector should be placed or cached in shared memory. Additionally, it can
control which vectors explicitly reside in registers by allocating space on the
stack frame, unrolling all inner loops of the kernel and substituting all
vector addressing with static values. The code generation strategy allows this
flexibility for the three vectors of concern, namely, $Probs_i$, $Grads_i$
and~$\pi_i$.
Hence, this strategy allows for 8 possible configurations denoted by
three letters each of which is a choice between $R$ or $S$ to represent
$Register$ or $Shared$ respectively. For example, $SSR$ denotes that $Probs_i$,
$Grads_i$ and $\pi_i$ were placed in $Shared$, $Shared$ and $Register$
respectively.

>>>>

A key structural change applied to the algorithm is the lossless compression of
its state. This modification enabled the algorithm to process larger data sets
while maintaining all program state in memory. Moreover, as the algorithm is
data-intensive, a reduction of the state is accompanied by a decrease in its
data intensity.
%
The data structures that occupy most memory are $\pi$ and~$\phi$, both of
which are matrices of dimension ${N}\times{K}$. However, storing both
is redundant as $\pi$ is a row-normalized copy of~$\phi$, see
Table~\ref{tbl-symbols} in Section~\ref{sec-background}. The full storage for
$\phi$ is discarded; $\phi_{ik}$ values are calculated as
$\pi_{ik}/\phi^{sum}_i$, which requires maintenance of a vector
$\phi^{sum}$ of size $N$.
Moreover, the calculation for $\phi_{ik}$ can be cached usefully.
%
The $\phi$ matrix is required in two kernels only, namely, \textit{update\_phi} and
\textit{update\_pi}. Both kernels access $\phi_i$ only for vertices~$i$ in the
mini-batch~\Minibatch, so for each iteration, the calculated values for $\phi$ are cached
in a smaller temporary matrix of size
$M\times{K}$.
%
% In this context, the initialization of the algorithm's state is performed by
% normalizing the full $\pi$ matrix in-place without the need for an additional
% matrix. Moreover, the normalization denominator of each row is cached in
% $\phi^{sum}_{i}$.
%
% Thus each read-only access to $\phi_{i,j}$ can be substituted with
% $\pi_{i,j}\phi_{i}^{sum}$. Given that the value of $\phi_{i,j}$ is consistently
% used in conjunction with $\pi_{i,j}$ and that $\phi_{i}^{sum}$ is read once and
% reused multiple times, this
This transformation trades memory storage and
bandwidth for a minimal computation overhead.
%
% On the other hand, the newly computed $\phi$ values can be stored in the
% temporary matrix and later be normalized into their corresponding rows of
% $\pi$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Single floating point vs double
Another modification to the original algorithm is migrating all of its
computation from double to single precision floating-point, analogous to the change
for the baseline implementation in Section~\ref{sec-seq-performance}.
%
This change affects the algorithm in multiple ways.
It again halves the memory
footprint of the algorithm and frees registers.
Besides, on our GPUs it
reduces
the computation intensity besides the data intensity.

@@

% \subsubsection{Edge-centric and Vertex-centric kernels}
\subsubsection{Memory Hierarchy Optimizations}
\label{gpu-design}

The work decomposition of the kernels ensures that each thread performs
independent computation in order to avoid expensive synchronization operations.
Edge-centric kernels, which operate over mini-batch edges, perform computations
over every edge in parallel.  Similarly, vertex-centric kernels that operate
over vertices in the mini-batch exploit parallelism across the selected
vertices.
% vectorization
Additionally, the kernels were vectorized to decrease instruction overhead.

% The GPU implementation builds on top of the CPU work decomposition scheme.
% However, instead of having every thread perform independent computations, each
% block of threads shares the work associated with the single edge or vertex for
% edge-centric and vertex-centric kernels.

Since all of the kernels are data-bound, several memory organization strategies
of varying complexity were investigated to speed up the algorithm by
exploiting the GPU's memory hierarchy.
% Update_phi
As a case study, a discussion of the \textit{update\_phi} kernel is provided as it is
the predominant component of the algorithm.

@@

% Naive strategy
\paragraph*{\textbf{Naive strategy}} allocate temporary vectors in thread
local memory to store $Probs$ and $Grads$ which physically reside in
device memory. All memory accesses are coalesced to achieve the highest
possible bandwidth to and from device memory.

@@

% Register and shared memory using code generation and loop unrolling 
\paragraph*{\textbf{Code generation strategy}} dynamically generate the code of the kernel to
custom tailor its properties. For example, this strategy can control whether a
vector should be placed or cached in shared memory. Additionally, it can
control which vectors explicitly reside in registers by allocating space on the
stack frame, unrolling all inner loops of the kernel and substituting all
vector addressing with static values. The code generation strategy allows this
flexibility for the three vectors of concern, namely, $Probs_i$, $Grads_i$
and~$\pi_i$.
Hence, this strategy allows for 8 possible configurations denoted by
three letters, each of which is a choice between $R$ or $S$ to represent
$Register$ or $Shared$ respectively. For example, $SSR$ denotes that $Probs_i$,
$Grads_i$ and $\pi_i$ are placed in $Shared$, $Shared$ and $Register$
respectively.

### 2017-05-30 gpu-05-evaluation.tex
Tiny textual fixes

The following sections provide a detailed experimental performance analysis
of the optimizations described above.
%
We assess the trade-offs associated with each of the GPU optimization
strategies and their performance effects on four GPUs spanning two chip
architecture generations.
%
Finally, we present the overall outcome of accelerating the algorithm by
displaying its fast convergence over the largest sensible combination of
community network and number of communities, so that the algorithm data fit in the
memory of a GPU.

>>>>

The following sections provide a detailed experimental performance analysis
of the optimizations described above.
%
We assess the trade-offs associated with each of the GPU optimization
strategies and their performance effects on four GPUs spanning two chip
architecture generations.
%
Finally, we present the overall outcome of accelerating the algorithm by
displaying its convergence speed over the largest sensible combination of
community network and number of communities while keeping the algorithm data
in the memory of a GPU.
